On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-11-26 paper/paper.tex
tweaking the bib
Removed a lot of commented-out LaTeX

%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\RequirePackage{amsmath,amssymb}
% \RequirePackage{stmaryrd}
\documentclass[runningheads, a4paper]{llncs}
% \usepackage{biblatex}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tkz-graph}
\usepackage{stmaryrd}
\usepackage{wrapfig}
% \usepackage{amsmath}

% \lstset{basicstyle=\ttfamily}
\usetikzlibrary{shapes.geometric, arrows}

@@

% TODO: Use these:
\newcommand{\exprt}[1]{\expr{\ttt{#1}}}
\newcommand{\tdots}{\ttt{...}}
\newcommand{\exprdots}{\expr{\tdots}}

@@

% \newcommand{\showtodos}{}  % Comment out to hide todos

\newenvironment{todo}
  {\ifthenelse{\isundefined{\showtodos}}{\comment}{\begin{tcolorbox}
    \textbf{TODO}:}}
  {\ifthenelse{\isundefined{\showtodos}}{\endcomment}{\end{tcolorbox}}
  }
\newenvironment{todont}
               {\comment}
               {\endcomment}

% \newenvironment<>{todo}[1][]{
%     \setbeamercolor{block body example}{fg=black,bg=blue}
%     \setbeamercolor{block title example}{fg=white,bg=red!75!black}
%     \setbeamertemplate{blocks}[rounded][shadow=false]
%   \begin{todo}[]}{\end{todo}
% }

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.

% \setcopyright{acmcopyright}
% \copyrightyear{2020}
% \acmYear{2020}
% \acmDOI{XX.XXXX/XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.

% \acmConference[GPCE '20]{GPCE '20: Proceedings of the 13th
%   ACM SIGPLAN International Symposium on Haskell}{November 15 -- 20, 2020}{Chicago, Illinois, United States}
% \acmBooktitle{Proceedings of the 19th ACM SIGPLAN International
% Conference on Generative Programming: Concepts and Experiences (GPCE '20)
%   November 15 -- 20, 2020, Chicago, Illinois, United States}
% \acmPrice{15.00}
% \acmISBN{XXX-X-XXXX-XXXX-X/XX/XX}

% \citestyle{acmauthoryear}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{On Adding Pattern Matching to Haskell-based
  Deeply Embedded Domain Specific Languages}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{David Young\inst{1} \and Mark Grebe\inst{2} \and Andy Gill\inst{1}}
\institute{University of Kansas \and University of Central Missouri}

\maketitle
% \authornote{Both authors contributed equally to this research.}
% \email{trovato@corporation.com}
% \orcid{1234-5678-9012}

% \author{Mark Grebe}
% \email{grebe@ucmo.edu}

% \author{Andrew Gill}
% \email{andygill@ku.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
% \renewcommand{\shortauthors}{Trovato and Tobin, et al.}

% \renewcommand{\shortauthors}{Young, Grebe and Gill}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.

% \linenumbers

@@

% \begin{CCSXML}
% <ccs2012>
%    <concept>
%        <concept_id>10011007.10011006.10011041.10011046</concept_id>
%        <concept_desc>Software and its engineering~Translator writing systems and compiler generators</concept_desc>
%        <concept_significance>300</concept_significance>
%        </concept>
%    <concept>
%        <concept_id>10011007.10011006.10011041.10011047</concept_id>
%        <concept_desc>Software and its engineering~Source code generation</concept_desc>
%        <concept_significance>300</concept_significance>
%        </concept>
%    <concept>
%        <concept_id>10011007.10011006.10011050.10011017</concept_id>
%        <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
%        <concept_significance>500</concept_significance>
%        </concept>
%  </ccs2012>
% \end{CCSXML}

% \ccsdesc[300]{Software and its engineering~Translator writing systems and compiler generators}
% \ccsdesc[300]{Software and its engineering~Source code generation}
% \ccsdesc[500]{Software and its engineering~Domain specific languages}

% \begin{CCSXML}
% <ccs2012>
%    <concept>
%        <concept_id>10011007.10011006.10011050.10011017</concept_id>
%        <concept_desc>Software and its engineering~Domain specific languages</concept_desc>
%        <concept_significance>500</concept_significance>
%        </concept>
%  </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~Domain specific languages}

% \keywords{domain specific languages}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
% \keywords{datasets, neural networks, gaze detection, text tagging}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.

@@

% \begin{todo}
%   Maybe add brief summary of relationship to lambdas?
% \end{todo}

@@

\begin{itemize}
  \item A representation of pattern matching in a Haskell EDSL (Section~\ref{sec:PatRep})
  \item A representation of lambdas (Section~\ref{sec:LamRep}) in the EDSL, utilizing
  the fact that both pattern matching and tail recursion are already captured in
  the EDSL.
  \item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section~\ref{sec:CorePlugin})
  \item An implementation of an interpreter for a reference semantics of the EDSL (Section~\ref{sec:StdSemantics}).
% \item An outline of a C backend for this EDSL (Section~\ref{sec:CBackend}).
\end{itemize}

% In this paper, a C backend is given as an example in Section \ref{sec:CBackend}. The Core plugin will
% translate a subset of Haskell to the EDSL we have specified. The C backend will
% then take expressions written in this EDSL and translate them to C code. This C
% code can be compiled by a standard C compiler. In effect, the subset of Haskell
% represented in the EDSL is compiled to C.

@@

\begin{todont}
  Expand?
\end{todont}

@@

% \newpage
Haskell EDSLs that support powerful language features like pattern matching
allow these advantages to be brought to problem domains where it is currently
either difficult or impossible to use a language that has these benefits.
However, there is no support for capturing pattern matching builtin to the Haskell
language itself. As a result, EDLs which capture pattern matching must do so
via a plugin in a Haskell compiler such as GHC. For each such EDSL, this mechanism
has needed to be reimplemented each time. Additionally, the task of implementing this
is difficult and error-prone. In this paper, we descibe a \textit{reusable}
framework for capturing pattern matching, as well as lambdas and tail recursion.

% A C backend is described in Section \ref{sec:CBackend}. However, other backends
% could be constructed, targeting different architectures.

\begin{todont}
  % Cite examples where Haskell implementations have been derived specifications.  Conal Elliott's latest two papers (as of May 2020), \cite{Elliott-2018-ad-icfp} and \cite{Elliott2019-convolution-extended}, would be good examples of this. (This is now cited)

  A paper which describes a Haskell implementation derived from an operational
  semantics would also be a good example.
\end{todont}

@@

% \begin{todo}
%   Does it make sense for this to be a subsection of the intro?
% \end{todo}

@@

% \begin{todo}
%   Look into \ttt{generic-sop} and compare then cite corresponding paper.
% \end{todo}

@@

% \begin{todo}
%   Should we make a synonym for this constraint since it is used in
% multiple places? Like this:
% \begin{lstlisting}
% type CanonicalIsIdem t =
%   ERepTy (ERepTy t) ~ ERepTy t
% \end{lstlisting}
% \end{todo}

@@

% \newpage
\begin{lstlisting}
data E t where
  ...
  SumMatchExp :: (ERep a, ERep b, ERepTy b ~ b) =>
    E (ProdMatch a r) -> E (SumMatch b r)
      -> E (SumMatch (Either a b) r)

@@

% \subsection{Connection to Church encodings}
% \label{sec:Church}
% % \begin{todo}
% %   Work on this subsection. Maybe this shouldn't be a separate section at all and
% %   maybe its contents should be incorporated throughout the rest of the paper.
% % \end{todo}

% Algebraic datatypes can be represented as lambda calculus terms using
% \textit{Church encodings}.~\cite{Boehm:1985:AutoSynth} For example, the pair constructor can be
% represented as

% \begin{lstlisting}
% church_mkPair
%   :: a -> b -> (a -> b -> r) -> r
% church_mkPair x y = \ f -> f x y
% \end{lstlisting}

% Coproducts can be represented with the constructors

% \begin{lstlisting}
% church_mkLeft
%   :: a -> (a -> r) -> (b -> r) -> r
% church_mkLeft x = \ f g -> f x

% church_mkRight
%   :: b -> (a -> r) -> (b -> r) -> r
% church_mkRight y = \ f g -> g y
% \end{lstlisting}

% In this representation:

% \begin{itemize}
%   \item A pair of type \ttt{A} and type \ttt{B} is represented with a value of type
% \ttt{forall r. (A -> B -> r) -> r}.

%   \item A coproduct of type \ttt{A} and type \ttt{B} is represented with a value of type
% \ttt{forall r. (A -> r) -> (B -> r) -> r}.
% \end{itemize}

% Both of those representations represent the type as a function that performs the
% corresponding case analysis. A deep embedding of the operations
% \ttt{churchMkPair}, \ttt{church\_mkLeft} and \ttt{church\_mkRight} is given
% by the type

% \begin{lstlisting}
% data Church r where
%   ChurchPair
%     :: a -> b -> (a -> b -> r) ->
%        Church r
%   ChurchLeft
%     :: a -> (a -> r) -> (b -> r) ->
%        Church r
%   ChurchRight
%     :: b -> (a -> r) -> (b -> r) ->
%        Church r
% \end{lstlisting}

% with the evaluation function

% \begin{lstlisting}
% churchEval :: Church r -> r
% churchEval (ChurchPair x y f) = f x y
% churchEval (ChurchLeft x f g) = f x
% churchEval (ChurchRight y f g) = g y
% \end{lstlisting}

% However, we only want to keep track of the functions used for elimination. By moving
% the values actually being contained in the products and coproducts out to the
% evaluation function, we arrive at

% \begin{lstlisting}
% data Church' t r where
%   ChurchPair'
%     :: (a -> b -> r) ->
%        Church' (a, b) r

%   ChurchEither'
%     :: (a -> r) -> (b -> r) ->
%        Church' (Either a b) r

% churchEval' :: Church' t r -> t -> r
% churchEval' (ChurchPair' f) (x, y) = f x y
% churchEval' (ChurchEither' f g) e =
%   case e of
%     Left  x -> f x
%     Right y -> g y
% \end{lstlisting}

% The \verb|Church'| type forms a large part of the foundation of the pattern
% matching aspect of the \verb|E| type. A rough correspondence between the two
% is given in the table below.\\

% {\small
% \noindent\begin{tabular}{|p{3.4cm}|p{4.4cm}|}
%   \hline
%   \ttt{Church' (a, b) r} & \ttt{E (SumMatch (a, b) r} \\
%   \hline
%   \ttt{Church' (Either a b) r} & \ttt{E (ProdMatch (Either a b) r)}\\
%   \hline
% \end{tabular}
% }

\begin{todo}
  Figure out how recursive types fit into this framework. Church encodings can
  handle recursive types and so can our representation. This is also the case
  for Scott encodings, etc, and is the difference between Church and Scott
  encodings.

  Does our way of handling recursive types correspond to how any of
  those encodings handle recursive types?

  % Maybe this is relevant: \href{https://www.reddit.com/r/haskell/comments/hvlh6a/generalized_church_encoding_is_the_curryhoward_of/}{\color{blue}{Generalized Church encoding is the Curry-Howard of Knaster-Tarski}}
\end{todo}

% \begin{lstlisting}
% data ChurchMatch t where
%   ChurchCoproductMatch
%     :: ChurchMatch ((a, b) -> r) -> ChurchMatch (Either b c -> r) ->
%        ChurchCoproductMatch (Either
% \end{lstlisting}

% \begin{todo}
%   Maybe it could make sense to essentially derive the pattern matching representation
%   from this Church encoding. If so, this could be described around here.
% \end{todo}

@@

% \begin{todo}
%   \ttt{runIter} has the following alternate implementation. Would this be
%   useful to use somewhere?

%   \begin{lstlisting}[otherkeywords={}]
%   runIter f = getDone . fix (f >=>)

%   getDone :: Iter b a -> b
%   getDone (Done x) = x
%   getDone _ = error "getDone"
%   \end{lstlisting}

%   where \verb|Iter| is given a standard \verb|Monad| instance (essentially the
%   same as the \verb|Either| \verb|Monad| instance). \\

%   Maybe this could be useful in some proofs or derivations somewhere (in this
%   paper or elsewhere)?

%   Note that \verb|error| is never called under any circumstance in this
%   implementation of \verb|runIter|.
% \end{todo}

@@

% We translate this to the expression language, and then apply two of
% the constructors we have defined.  The first of these is an initial 
% \ttt{Done} which we then wrap with a \ttt{runIter} to provide the
% representation of the iteration execution.
% The transformation then proceeds by pushing the DoneExp constructor through
% the conditionals, and the recursive call is transformed into a StepExp constructor.
% This leaves us the the transformed example:

@@

% \break

@@

% \newpage

% \section{C Backend}
% \label{sec:CBackend}

% \begin{todont}
%   Should this section be expanded?
% \end{todont}

% The C backend makes use of a small EDSL for generating C code. See
% Appendix B for implementation details of this ESDL.

% The following \ttt{struct}s provide the C representation for Haskell expressions:

% \begin{lstlisting}
% typedef enum var_type_tag {
%   EXPR_INT
% , EXPR_PAIR
% ...
% , EXPR_UNIT
% } var_type_tag;

% typedef enum semantic_type_tag {
%   NO_SEMANTIC_TAG
% , EXPR_COMPLEX     // Complex numbers
% } semantic_type_tag;

% struct closure_t;

% typedef struct var_t {
%   var_type_tag tag;
%   semantic_type_tag semantic_tag;
%   void* value;
% } var_t;

% typedef struct closure_t {
%   var_t* fv_env;
%   var_t (*fn)(var_t, struct closure_t*);
% } closure_t;
% \end{lstlisting}

% The \ttt{semantic\_tag} allows type information preserved by the \ttt{Typeable} constraints
% in the EDSL to be brought over to the generated C code. In our implementation, this is
% used to distinguish between pairs of \ttt{Double}s and \ttt{Complex Double} values. In the
% case of a \ttt{Complex Double}, the \ttt{var\_t} would have a \ttt{tag} field with the value
% \ttt{EXPR\_PAIR} and a \ttt{semantic\_tag} field with the value \ttt{EXPR\_COMPLEX}.

% A C code generation function \ttt{genC} is implemented:

% \begin{lstlisting}
% genC :: E a -> String
% ...
% \end{lstlisting}

% To use this, for example, with the previously mentioned \ttt{isEven} implementation we
% can write:
% % \newpage
% \begin{lstlisting}
% isEvenE :: E (Int -> Bool)
% isEvenE =
%   externalize (
%     case x == 0 of
%       True -> True
%       False ->
%         case x == 1 of
%           True -> False
%           False -> isEvenE (x - 2))

% isEvenC :: String
% isEvenC = genC isEvenE
% \end{lstlisting}

@@

The plugin's control flow is described in Figure~\ref{fig:CorePlugin}. When control
flow reaches a node, the rewrite rules listed in parentheses are performed. These rewrite
rule names correspond to the rules given in Figure~\ref{fig:Rewrites}.
% In the flowchart given in Figure~\ref{fig:CorePlugin}, the names in parentheses
% in a node refer to the names of transformations given in the rewrite rules in
% Figure~\ref{fig:Rewrites}. The control flow of the plugin is described by the
% flowchart and the rewrites performed when control reaches a node is given by the list of
% rule names in parentheses in that node.
In Figure~\ref{fig:Rewrites},
$\varepsilon$ represents an empty term (in the context of the term rewriting system used in that Figure). Calls
to \ttt{unrep :: E a -> a} are only used internally and will not exist in
the final result of the transformation.

% \begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}
  \centering
  % \vspace{-1cm}
  \resizebox{0.8\columnwidth}{8.2cm}{\includegraphics{Diagrams/Flowchart.pdf}}
  % \includegraphics{Diagrams/Flowchart.pdf}
   \caption{Control Flow of the Core Plugin}
   \label{fig:CorePlugin}
  % \vspace{-0.1cm}
\end{figure}%

@@

% \clearpage

% \clearpage

% \subsection{Syntactic Transformations of the Core Plugin}

% \begin{todo}
%   Should we use Oxford brackets here? Maybe a bit unusual, but we are operating syntactically and mapping from \ttt{a} to \ttt{E a} (could \ttt{E a} be considered an "intermediary" semantic domain?).
% \end{todo}
% \vspace*{-0.6cm}

% \begin{wrapfigure}{r}{0.8\linewidth}
  % \begin{wrapfigure}{r}{0.59\linewidth}
\begin{figure}
  \centering
  % \vspace{-0.9cm}
\resizebox{1\linewidth}{!}{
  \begin{minipage}{\linewidth}
\begin{align*}
  &D\expr{\ttt{f x = internalize (externalize (case s of \{ ... \}))}}\\
      &\quad\rewrites I(D(\ttt{f x = externalize (case s of \{ ... \})}))\\
  &D\expr{\ttt{f x = externalize (case s of \{ ... \})}}\\
      &\quad\rewrites \begin{cases}
        \ttt{f = $C^{*}\expr{\ttt{externalize ($\lambda \eta \rarr$ runIter ($\lambda \ttt{x} \rarr \ttt{case s of \{ $T'_{\ttt{f}}\exprdots$}$ \}) $\eta$)}}$} & \text{\parbox{2.4cm}{if \ttt{f} occurs free in \ttt{\{ ... \}}, where $\eta$ is a fresh name}}\\
        \\
        \ttt{f x = $C^{*}\expr{\ttt{externalize (case s of \{ ... \})}}$} & \text{otherwise}
      \end{cases}\\
  &\\
  &I\expr{\ttt{f x = externalize e}} \rewrites \ttt{f x = internalize (externalize e)}\\
  &I\expr{\ttt{f = externalize e}} \rewrites \ttt{f = internalize (externalize e)}\\
  &T'_{\ttt{f}}\expr{\ttt{K x0 ...$_U$ xN $\rarr$ case s of \{ ...$_V$ \}; ...$_W$}}\\
    &\quad\rewrites
        \ttt{K x0 ...$_U$ xN $\rarr$ $T_{\ttt{f}}\expr{\ttt{case s of \{ ...$_V$ \}}}$; $T'_{\ttt{f}}\expr{\ttt{ ...$_W$ }}$}\\
  &T'_{\ttt{f}}\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$ }}\\
    &\quad\rewrites
      \begin{cases}
        \ttt{K x0 ...$_U$ xN $\rarr$ body0}[\ttt{f} \mapsto \ttt{Step}]\ttt{; }T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }} & \text{if \ttt{f} occurs free in \ttt{body0}}\\
        \ttt{K x0 ...$_U$ xN $\rarr$ Done body0; $T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }}$} & \text{otherwise}
      \end{cases}\\
  &T'_{\ttt{f}}\expr{\varepsilon} \rewrites \varepsilon\\
  &T_{\ttt{f}}\expr{\ttt{case s of \{ ... \}}}\\
    &\quad\rewrites
      \begin{cases}
        \ttt{case s of \{ $T'_{\ttt{f}}\expr{\ttt{ ... }}$ \}} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{Done (case s of \{ ... \})} & \text{otherwise}
      \end{cases}\\
  &C^{*}\expr{\ttt{x}} \rewrites
    \begin{cases}
      \ttt{x} & \text{if \ttt{x} has no subexpressions marked with \ttt{externalize}}\\
      C^{*}\expr{C\expr{\ttt{y}}} & \text{if \ttt{externalize y} is the first marked subexpression of \ttt{x}}\\
    \end{cases}\\
  &C\expr{\ttt{runIter f x}} \rewrites \ttt{App (TailRec $M\expr{\ttt{f}}$) $M\expr{\ttt{x}}$}\\
  &C\expr{\ttt{x + y}} \rewrites \ttt{Add $M\expr{\ttt{x}}$ $M\expr{\ttt{y}}$}\\
  &C\expr{\ttt{case scrutinee of \{ ... \}}} \rewrites \ttt{CaseExp $M\expr{\ttt{scrutinee}}$ } B\expr{\ttt{ ... }}\\
  &C\expr{\ttt{$\lambda$x $\rarr$ body}} \rewrites A\expr{\lambda \ttt{x} \rarr \ttt{body}}\\
  &C\expr{\ttt{K x0 ... xN }}\\
    &\quad\rewrites \ttt{construct (K (unrep $M\expr{\ttt{x0}}$) ... (unrep $M\expr{\ttt{xN}}$))}\;\;\;\;\text{(Where \ttt{K} is a constructor)}\\
  &C\expr{\ttt{f x}} \rewrites \ttt{App $M\expr{f}$ $M\expr{x}$}\\
  % &C\expr{\ttt{f x}} \rewrites \ttt{ConstructAp $M\expr{\ttt{f}}$ $M\expr{\ttt{x}}$}\;\;\;\;\text{(Where \ttt{f} is not a primitive or an application of a primitive)}\\
  % &C\expr{\ttt{x}} \rewrites \ttt{Construct x}\;\;\;\;\text{(Where \ttt{x} is a variable which is not a primitive)}\\
  &B\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$}} \rewrites \ttt{SumMatchExp }P\expr{\ttt{x0 ...$_U$ xN $\rarr$ body0}}\; B\expr{\ttt{ ...$_V$ }}\\
  &B\expr{\ttt{K x0 ... xN $\rarr$ body}} \rewrites \ttt{OneSumMatchExp }P\expr{\ttt{x0 ... xN $\rarr$ body}}\\
  &B\expr{\ttt{$\varepsilon$}} \rewrites \ttt{EmptyMatch}\\
  &P\expr{\ttt{x0 x1 ... xN $\rarr$ body}} \rewrites \ttt{ProdMatchExp }A\expr{\ttt{$\lambda$ x0 $\rarr$ $P\expr{\ttt{x1 ... xN $\rarr$ body}}$}}\\

>>>>

\RequirePackage{amsmath,amssymb}

\documentclass[runningheads, a4paper]{llncs}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tkz-graph}
\usepackage{stmaryrd}
\usepackage{wrapfig}

\usetikzlibrary{shapes.geometric, arrows}

@@

% TODO: Use these:

\newcommand{\exprt}[1]{\expr{\ttt{#1}}}
\newcommand{\tdots}{\ttt{...}}
\newcommand{\exprdots}{\expr{\tdots}}

@@

\newenvironment{todo}
  {\ifthenelse{\isundefined{\showtodos}}{\comment}{\begin{tcolorbox}
    \textbf{TODO}:}}
  {\ifthenelse{\isundefined{\showtodos}}{\endcomment}{\end{tcolorbox}}
  }
\newenvironment{todont}
               {\comment}
               {\endcomment}
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

\title{On Adding Pattern Matching to Haskell-based
  Deeply Embedded Domain Specific Languages}

@@

\author{David Young\inst{1} \and Mark Grebe\inst{2} \and Andy Gill\inst{1}}
\institute{University of Kansas \and University of Central Missouri}

\maketitle

@@

@@

@@

\begin{itemize}
  \item A representation of pattern matching in a Haskell EDSL (Section~\ref{sec:PatRep})
  \item A representation of lambdas (Section~\ref{sec:LamRep}) in the EDSL, utilizing
  the fact that both pattern matching and tail recursion are already captured in
  the EDSL.
  \item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section~\ref{sec:CorePlugin})
  \item An implementation of an interpreter for a reference semantics of the EDSL (Section~\ref{sec:StdSemantics}).
\end{itemize}

@@

@@

Haskell EDSLs that support powerful language features like pattern matching
allow these advantages to be brought to problem domains where it is currently
either difficult or impossible to use a language that has these benefits.
However, there is no support for capturing pattern matching builtin to the Haskell
language itself. As a result, EDLs which capture pattern matching must do so
via a plugin in a Haskell compiler such as GHC. For each such EDSL, this mechanism
has needed to be reimplemented each time. Additionally, the task of implementing this
is difficult and error-prone. In this paper, we descibe a \textit{reusable}
framework for capturing pattern matching, as well as lambdas and tail recursion.

@@

@@

@@

@@

\begin{lstlisting}
data E t where
  ...
  SumMatchExp :: (ERep a, ERep b, ERepTy b ~ b) =>
    E (ProdMatch a r) -> E (SumMatch b r)
      -> E (SumMatch (Either a b) r)

@@

@@

@@

@@

@@

@@

The plugin's control flow is described in Figure~\ref{fig:CorePlugin}. When control
flow reaches a node, the rewrite rules listed in parentheses are performed. These rewrite
rule names correspond to the rules given in Figure~\ref{fig:Rewrites}.
%
In Figure~\ref{fig:Rewrites},
$\varepsilon$ represents an empty term (in the context of the term rewriting system used in that Figure). Calls
to \ttt{unrep :: E a -> a} are only used internally and will not exist in
the final result of the transformation.

\begin{figure}
  \centering
  \resizebox{0.8\columnwidth}{8.2cm}{\includegraphics{Diagrams/Flowchart.pdf}}
   \caption{Control Flow of the Core Plugin}
   \label{fig:CorePlugin}
\end{figure}%

@@

\begin{figure}
  \centering
\resizebox{1\linewidth}{!}{
  \begin{minipage}{\linewidth}
\begin{align*}
  &D\expr{\ttt{f x = internalize (externalize (case s of \{ ... \}))}}\\
      &\quad\rewrites I(D(\ttt{f x = externalize (case s of \{ ... \})}))\\
  &D\expr{\ttt{f x = externalize (case s of \{ ... \})}}\\
      &\quad\rewrites \begin{cases}
        \ttt{f = $C^{*}\expr{\ttt{externalize ($\lambda \eta \rarr$ runIter ($\lambda \ttt{x} \rarr \ttt{case s of \{ $T'_{\ttt{f}}\exprdots$}$ \}) $\eta$)}}$} & \text{\parbox{2.4cm}{if \ttt{f} occurs free in \ttt{\{ ... \}}, where $\eta$ is a fresh name}}\\
        \\
        \ttt{f x = $C^{*}\expr{\ttt{externalize (case s of \{ ... \})}}$} & \text{otherwise}
      \end{cases}\\
  &\\
  &I\expr{\ttt{f x = externalize e}} \rewrites \ttt{f x = internalize (externalize e)}\\
  &I\expr{\ttt{f = externalize e}} \rewrites \ttt{f = internalize (externalize e)}\\
  &T'_{\ttt{f}}\expr{\ttt{K x0 ...$_U$ xN $\rarr$ case s of \{ ...$_V$ \}; ...$_W$}}\\
    &\quad\rewrites
        \ttt{K x0 ...$_U$ xN $\rarr$ $T_{\ttt{f}}\expr{\ttt{case s of \{ ...$_V$ \}}}$; $T'_{\ttt{f}}\expr{\ttt{ ...$_W$ }}$}\\
  &T'_{\ttt{f}}\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$ }}\\
    &\quad\rewrites
      \begin{cases}
        \ttt{K x0 ...$_U$ xN $\rarr$ body0}[\ttt{f} \mapsto \ttt{Step}]\ttt{; }T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }} & \text{if \ttt{f} occurs free in \ttt{body0}}\\
        \ttt{K x0 ...$_U$ xN $\rarr$ Done body0; $T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }}$} & \text{otherwise}
      \end{cases}\\
  &T'_{\ttt{f}}\expr{\varepsilon} \rewrites \varepsilon\\
  &T_{\ttt{f}}\expr{\ttt{case s of \{ ... \}}}\\
    &\quad\rewrites
      \begin{cases}
        \ttt{case s of \{ $T'_{\ttt{f}}\expr{\ttt{ ... }}$ \}} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{Done (case s of \{ ... \})} & \text{otherwise}
      \end{cases}\\
  &C^{*}\expr{\ttt{x}} \rewrites
    \begin{cases}
      \ttt{x} & \text{if \ttt{x} has no subexpressions marked with \ttt{externalize}}\\
      C^{*}\expr{C\expr{\ttt{y}}} & \text{if \ttt{externalize y} is the first marked subexpression of \ttt{x}}\\
    \end{cases}\\
  &C\expr{\ttt{runIter f x}} \rewrites \ttt{App (TailRec $M\expr{\ttt{f}}$) $M\expr{\ttt{x}}$}\\
  &C\expr{\ttt{x + y}} \rewrites \ttt{Add $M\expr{\ttt{x}}$ $M\expr{\ttt{y}}$}\\
  &C\expr{\ttt{case scrutinee of \{ ... \}}} \rewrites \ttt{CaseExp $M\expr{\ttt{scrutinee}}$ } B\expr{\ttt{ ... }}\\
  &C\expr{\ttt{$\lambda$x $\rarr$ body}} \rewrites A\expr{\lambda \ttt{x} \rarr \ttt{body}}\\
  &C\expr{\ttt{K x0 ... xN }}\\
    &\quad\rewrites \ttt{construct (K (unrep $M\expr{\ttt{x0}}$) ... (unrep $M\expr{\ttt{xN}}$))}\;\;\;\;\text{(Where \ttt{K} is a constructor)}\\
  &C\expr{\ttt{f x}} \rewrites \ttt{App $M\expr{f}$ $M\expr{x}$}\\
  &B\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$}} \rewrites \ttt{SumMatchExp }P\expr{\ttt{x0 ...$_U$ xN $\rarr$ body0}}\; B\expr{\ttt{ ...$_V$ }}\\
  &B\expr{\ttt{K x0 ... xN $\rarr$ body}} \rewrites \ttt{OneSumMatchExp }P\expr{\ttt{x0 ... xN $\rarr$ body}}\\
  &B\expr{\ttt{$\varepsilon$}} \rewrites \ttt{EmptyMatch}\\
  &P\expr{\ttt{x0 x1 ... xN $\rarr$ body}} \rewrites \ttt{ProdMatchExp }A\expr{\ttt{$\lambda$ x0 $\rarr$ $P\expr{\ttt{x1 ... xN $\rarr$ body}}$}}\\
  &P\expr{\ttt{x $\rarr$ body}} \rewrites \ttt{OneProdMatchExp }A\expr{\ttt{$\lambda$ x $\rarr$ body}}\\
  &P\expr{\ttt{ $\rarr$ body}} \rewrites \ttt{NullaryMatch $M\expr{\ttt{body}}$}\\
  &A\expr{\ttt{$\lambda$(x :: a) $\rarr$ body}} \rewrites \ttt{Lam (Name @a uniq) $M\expr{\ttt{body$[\ttt{x} \mapsto \ttt{unrep (Var @a uniq)}]$}}$}\\
  &\text{\;\;\;\;\;(where \ttt{uniq} is a globally unique identifier)}\\
  &M\expr{\ttt{unrep x}} \rewrites \ttt{x}\\
  &M\expr{\ttt{x :: a}} \rewrites
    \begin{cases}
      \ttt{externalize x} & \text{if}\; \nexists t, a\; \typeeq E\; t\\
      \ttt{x} & \text{if}\; \exists t, a\; \typeeq E\; t
    \end{cases}
\end{align*}
  \end{minipage}
}%

\caption{Rewrite rules}
\label{fig:Rewrites}
